Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-12-04 draft.tex
Minor changes at Formula
removed blind status

% TODO: BLIND Version
\author{Blind}
% \author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
% \affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
% \affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
% \affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
% \email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
% }

@@

%% TODO BLIND "We incorporate their" -> We incorporate the ...
We incorporate their IVA approach by concurrently displaying complex relationships between variables via decision trees.

@@

\paragraph{The Error Term. }
Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
%%
The metric usually works well for variables with uniform distribution.
%%
It distorts the result for other distribution types.
%%
For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, the error metric would yield an error of 10\%, even though it is very bad.
%%
Based on this we use a summary error, which incorporates the discriminative power of each manifestation and is denoted as follows:
%Our error term therefore incorporates the discriminative power of each manifestation and is denoted as follows:
%%
\begin{equation}
error = 1 - \frac{\sum_{m=1}^M \frac{correctlyClassified_{m}}{M}}{number_{dimensions}}
\end{equation}
$M$ represents the set of manifestation of each variable.
%%
The error is scaled to denote perfect classification with 0; 1 is equal to random selection.
%%
We consider a result rated $0.5$ or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
%%
It allows for comparability of error rates between variables with different manifestation count.

@@

The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} or \emph{curvature angle} with an error between $0.35$ and $0.47$.
%%
\subsection{Size Groups}
%%
Many previously described results are influenced by \emph{subject size}.
%%
Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is whether the inter-group spine variability parameter is sufficient for predicting other variables or not.
%%
Dividing subjects into \emph{body size} groups potentially highlights classifications not influenced by \emph{body size}.
%%
\paragraph{Large Decision Trees. }
Back pain-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
%%
\emph{Radiating back pain} could be described with an error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} with \emph{torsion} and \emph{mean curvature}.
%%
For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
%%
There are several decision trees for laboratory values, e.g. \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
%%
Similar values can be observed for \emph{cholesterol} or \emph{age}.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with an error between $0.42$ and $0.47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
%%
%TODO: Muss das rein? Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)
%%
\section{\uppercase{Conclusion}}
\label{sec:Conclusion}
\noindent The presented results indicate that the image-based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not sufficient to characterize pathological changes in the lumbar spine in the given quality.
%%
They are associated with subject \emph{gender} due to their \emph{body size} difference using \emph{mean curvature} and \emph{mean curvature angle}.
%%
This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between male and female.
%%
\emph{Gender} was detected well in the analysis of all subjects as well as subjects divided by \emph{BMI}.

>>>>

% \author{Blind}
\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
\affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
\affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
\email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
}

@@

We incorporate the IVA approach by concurrently displaying complex relationships between variables via decision trees.

@@

\paragraph{The Error Term. }
Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
%%
The metric usually works well for variables with uniform distribution.
%%
It distorts the result for other distribution types.
%%
For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, the error metric would yield an error of 10\%, even though it is very bad.
%%
Based on this we use a summary error, which incorporates the discriminative power of each manifestation and is denoted as follows:
%Our error term therefore incorporates the discriminative power of each manifestation and is denoted as follows:
%%
\begin{equation}
totalError = 1 - \sum_{m=1}^M \frac{correctlyClassified_{m}}{M\cdot N_m}
\end{equation}
$M$ represents the set of manifestation of each variable.
%%
The error is scaled to denote perfect classification with 0; 1 is equal to random selection.
%%
We consider a result rated $0.5$ or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
%%
It allows for comparability of error rates between variables with different manifestation count.

@@

The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} or \emph{curvature angle} with an error between $0.35$ and $0.47$.
%%
\subsection{Size Groups}
%%
Many previously described results are influenced by \emph{subject size}.
%%
Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is whether the inter-group spine variability parameter is sufficient for predicting other variables or not.
%%
Dividing subjects into \emph{body size} groups potentially highlights classifications not influenced by \emph{body size}.
%%
\paragraph{Large Decision Trees. }
Back pain-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
%%
\emph{Radiating back pain} could be described with an error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} with \emph{torsion} and \emph{mean curvature}.
%%
For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
%%
There are several decision trees for laboratory values, e.g. \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
%%
Similar values can be observed for \emph{cholesterol} or \emph{age}.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with an error between $0.42$ and $0.47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
%%
%TODO: [medium] Muss das rein? Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)
%%
\section{\uppercase{Conclusion}}
\label{sec:Conclusion}
\noindent The presented results indicate that the image-based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not sufficient to characterize pathological changes in the lumbar spine in the given quality.
%%
They are associated with subject \emph{gender} due to their \emph{body size} difference using \emph{mean curvature} and \emph{mean curvature angle}.
%%
This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between male and female.
%%
\emph{Gender} was detected well in the analysis of all subjects as well as subjects divided by \emph{BMI}.
